---
title: "Exploratory analysis with DESeq2"
author: "You!"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='./')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes  
  html_document:
    theme: yeti # nice theme for the webpage
    toc: yes # table of contents
    toc_float: yes # table of contents "floats" in the document
    df_print: paged # data frames are interactive
    dev: png # what format do you want for the figures?
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
DOCNAME = gsub(DOCNAME, pattern = ".Rmd", replacement = "", fixed = T)
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("./img/", DOCNAME, "/"), #images will be put in this folder, under the notebook name
                      message        = FALSE,
                      warning        = FALSE,
                      eval           = TRUE)
```

```{r, include = FALSE, echo = FALSE}
# DO NOT RUN IF YOU HAVE ALREADY RUN PREVIOUS NOTEBOOKS
# This chunk is ONLY necessary if you want to knit this document into a pdf!!
library(tidyverse)
library(DESeq2)
library(tximport)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
meta <- read_csv("../Data/samplesheet.csv")
meta$condition = factor(meta$condition, levels = c("vampirium", "control", "garlicum"))
dir <- "../Data/salmon"
tx2gene <- read_table("../Data/salmon/salmon_tx2gene.tsv", col_names = c("transcript_ID","gene_ID","gene_symbol"))
files <- file.path(dir, meta$sample, "quant.sf")
names(files) <- meta$sample
txi <- tximport(files, type="salmon", tx2gene=tx2gene, countsFromAbundance = "lengthScaledTPM", ignoreTxVersion	= TRUE)
dds <- DESeqDataSetFromTximport(txi,
                                   colData = meta %>% column_to_rownames("sample"), 
                              design = ~ condition)
keep <- rowSums(counts(dds)) > 0
dds <- dds[keep,]
```

Approximate time: 40 minutes

## Learning Objectives 

* Recognize the importance of methods for count data transformation
* Describe the PCA (principal component analysis) technique
* Interpret different examples of PCA plots
* Evaluate sample quality using PCA and hierarchical clustering 

## Vampirium quality assessment and exploratory analysis using DESeq2	

Now that we have a good understanding of the QC steps normally employed for RNA-seq, let's implement them for the Vampirium dataset we are going to be working with.

### Transform normalized counts for the Vampirium dataset

**To improve the distances/clustering for the PCA and hierarchical clustering visualization methods**, we need to moderate the variance across the mean by applying the rlog transformation to the normalized counts. 

```{r}
### Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

## Or using the vst for > 20 samples
#vsd <- vst(dds, blind = TRUE)
```

The `blind=TRUE` argument is to make sure that the `rlog()` function does not take our sample groups into account - i.e. does the transformation in an unbiased manner. When performing quality assessment, it is important to include this option. 

The `rlog()` function returns a `DESeqTransform` object, another type of DESeq-specific object. The reason you don't just get a matrix of transformed values is because all of the parameters (i.e. size factors) that went into computing the rlog transform are stored in that object. We use this object to plot the PCA and hierarchical clustering figures for quality assessment.

### Principal component analysis (PCA) for the Vampirium dataset

We are now ready for the QC steps, let's start with PCA!

DESeq2 has a built-in function for generating PCA plots using `ggplot2` under the hood. In addition, it takes the `rlog` object as an input directly, hence saving us the trouble of extracting the relevant information from it.

The function `plotPCA()` requires two arguments as input: a `DESeqTransform` object and the "intgroup" (interesting group), i.e. the name of the column in our metadata that has information about the experimental sample groups. 

```{r DESeq_PCA}
### Plot PCA 
plotPCA(rld, intgroup="condition")
```

***

**Exercise 1**:

By default `plotPCA()` uses the *top 500 most variable genes*. You can change this by adding the `ntop=` argument and specifying how many of the genes you want the function to consider. For example, try 1000 genes. Did the plot change a lot?

**Your code here:**

```{r}

```

***

**Exercise 2**:

1. What does the above plot tell you about the similarity of samples?
2. Does it fit the expectation from the experimental design?
3. What do you think the %variance information (in the axes titles) tell you about the data in the context of the PCA?

***

**Extra**

The `plotPCA()` function will only return the values for PC1 and PC2. If you would like to explore the additional PCs in your data or if you would like to identify genes that contribute most to the PCs, you can use the `prcomp()` function. For example, to plot any of the PCs we could run the following code:

```{r}
# Input is a matrix of log transformed values
rld_mat <- assay(rld) # extract rlog count matrix
pca <- prcomp(t(rld_mat)) # perform PCA on the transposed (t) matrix of data 
```

To see what the PCA object contains we can use again the `attributes()` function:

```{r}
attributes(pca)
```

You can check the `?prcomp()` for more information. The most important variables are: 
- `sdev`: standard deviation explained by each PC.
- `rotation`: contribution of each gene to each PC.
- `x`: PC values for each sample (we use this values for our plots)

We can create a new object that contains all our metadata information and the PC values:

```{r}
df <- cbind(meta, pca$x) # Create data frame with metadata and PC3 and PC4 values for input to ggplot
```

```{r custom_PCA}
# ggplot with info for all PCAs
ggplot(df) + geom_point(aes(x=PC3, y=PC4, color = condition))
```

If you want to add PC variation information to the plot we can fetch it using the summary() function and take the second row:

```{r}
summary(pca)
pca_var <- summary(pca)$importance[2,] # second row is stored in the object "importance"
pca_var <- round(pca_var * 100, digits = 2) # make it percentage and round to 2 digits
```

Finally, we can add it to our plot:

```{r}
ggplot(df) + geom_point(aes(x=PC3, y=PC4, color = condition)) + 
  xlab(paste0("PC3: ",pca_var["PC3"], "% variance")) + 
  ylab(paste0("PC4: ",pca_var["PC4"], "% variance")) 
```

### Hierarchical Clustering for the Vampirium dataset

There is no built-in function in DESeq2 for plotting the heatmap for displaying the pairwise correlation or distances between all the samples and the hierarchical clustering information; we will use the `pheatmap()` function from the `pheatmap` package. This function cannot use the `DESeqTransform` object as input, but requires a matrix or dataframe. So, the first thing to do is retrieve that information from the `rld` object using a function called `assay()`.

```{r}
### Extract the rlog matrix from the object
rld_mat <- assay(rld)    
```

Next, we need to compute the distances values for all the samples. We can do this using the `dist` function:

```{r}
sampleDists <- dist(t(rld_mat)) #distances are computed by rows, so we need to transpose (t) the matrix
sampleDistMatrix <- as.matrix(sampleDists)
```

Let's take a look at the column and row names of the correlation matrix. 

```{r}
## check the output of sampleDistMatrix, make note of the row names and column names
head(sampleDistMatrix)

head(meta)
```

You will notice that they match the names we have given our samples in the metadata data frame we started with. It is important that these match, so we can use the `annotation` argument below to plot a color block across the top. This block enables easy visualization of the hierarchical clustering.

Now, let's plot the heatmap!

```{r dist_heatmap}
### Load pheatmap package
library(pheatmap)

pheatmap(sampleDistMatrix, annotation_col = meta %>% column_to_rownames("sample") %>% 
           dplyr::select(condition)) # we only want to use the condition column as an annotation
```

When you plot using `pheatmap()` the hierarchical clustering information is used to place similar samples together and this information is represented by the tree structure along the axes. The `annotation` argument accepts a dataframe as input, in our case it is the `meta` data frame. 

**Exercise 2**

Instead of using distances between expression patterns, check the Pearson correlation between samples using `cor()`. Use your rlog count matrix as an input.

**Your code here:**

```{r}

```

**Extra**

There are many arguments and options for the `pheatmap()` function. You could, for example, change the color scale used, remove the dendograms, avoid clustering or even scale the values per row or per column.

```{r}
library(RColorBrewer)
heat.colors <- brewer.pal(6, "Blues") # Colors from the RColorBrewer package (only 6)
heat.colors <- colorRampPalette(heat.colors)(100) # Interpolate 100 colors
```

```{r custom_heatmap}
pheatmap(sampleDistMatrix, annotation = meta %>% column_to_rownames("sample") %>% dplyr::select("condition"), 
         color = heat.colors, border_color=NA, fontsize = 10, 
         fontsize_row = 10, height=20)
```       

You can check all the colors that RColorBrewer offers by using the following command: `display.brewer.all()`

```{r RcolorBrewer_colors, fig.height=10}
display.brewer.all()
```

