---
title: "The RNAseq count matrix"
author: "You!"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='./')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
DOCNAME = gsub(DOCNAME, pattern = ".Rmd", replacement = "", fixed = T)
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("./img/", DOCNAME, "/"),
                      #fig.width      = 10,
                      #fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE,
                      eval           = FALSE)
```

Approximate time: 20 minutes

## Learning Objectives

-   Describe how to set up an RNA-seq project in R
-   Describe the RNA-seq and the differential gene expression analysis workflow
-   Explain why negative binomial distribution is used to model RNA-seq count data

### Loading libraries

For this analysis we will be using several R packages, some which have been installed from CRAN and others from Bioconductor. To use these packages (and the functions contained within them), we need to **load the libraries.** 

```{r}
library(tidyverse)
library(RColorBrewer)
library(DESeq2)
library(pheatmap)
library(DEGreport)
library(ggrepel)
```

### Loading data

To load the data into our current environment, we will be using the `read_table` function. We need to provide the path to each file. By default the function expects tab-delimited files, which is what we have.

```{r}
## Load in data
data <- read_table("/work/introduction_bulkRNAseq_analysis/Data/Mov10_full_counts.txt") 

meta <- read_table("/work/introduction_bulkRNAseq_analysis/Data/Mov10_full_meta.txt")
```

Use `class()` to inspect our data and make sure we are working with data frames:

```{r}
### Check classes of the data we just brought in
class(meta)
class(data)
```

### Viewing data

Make sure your datasets contain the expected samples / information before proceeding to perfom any type of analysis.

```{r}
View(meta)
View(data)
```

### RNA-seq count distribution

To determine the appropriate statistical model, we need information about the distribution of counts. To get an idea about how RNA-seq counts are distributed, let's plot the counts of all the samples: 

```{r}
pdata <- data %>% 
  gather(key = Sample, value = Count, -GeneSymbol)

pdata

ggplot(pdata) +
  geom_histogram(aes(x = Count), stat = "bin", bins = 200) +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```

If we zoom in close to zero, we can see a large number of genes with counts of zero:

```{r}
ggplot(pdata) +
   geom_histogram(aes(x = Count), stat = "bin", bins = 200) + 
   xlim(-5, 500)  +
   xlab("Raw expression counts") +
   ylab("Number of genes")
```


These images illustrate some common features of RNA-seq count data, including a **low number of counts associated with a large proportion of genes**, and a long right tail due to the **lack of any upper limit for expression**. Unlike microarray data, which has a dynamic range maximum limited due to when the probes max out, there is no limit of maximum expression for RNA-seq data. Due to the differences in these technologies, the statistical models used to fit the data are different between the two methods.

### Modeling count data

RNAseq count data can be modeled using a **Poisson distribution**. this particular distribution is fitting for data where the **number of cases is very large but the probability of an event occurring is very small**. To give you an example, think of the lottery: many people buy lottery tickets (high number of cases), but only very few win (the probability of the event is small).

With RNA-Seq data, **a very large number of RNAs are represented and the probability of pulling out a particular transcript is very small**. Thus, it would be an appropriate situation to use the Poisson distribution. However, a unique property of this distribution is that the mean == variance. Realistically, with RNA-Seq data there is always some biological variation present across the replicates (within a sample class). Genes with larger average expression levels will tend to have larger observed variances across replicates.

The model that fits best, given this type of variability observed for replicates, is the **Negative Binomial (NB) model**. Essentially, **the NB model is a good approximation for data where the mean \< variance**, as is the case with RNA-Seq count data.

Run the following code to plot the *mean versus variance* for the 'Mov10 overexpression' replicates:

```{r}
df <- data %>% 
  rowwise() %>% 
  summarise(mean_counts = mean(Mov10_kd_2:Irrel_kd_3), 
            variance_counts = var(Mov10_kd_2:Irrel_kd_3))


ggplot(df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) + 
  geom_abline(intercept = 0, slope = 1, color="red") +
  scale_y_log10() +
  scale_x_log10()
```
