---
title: "Count normalization with DESeq2"
author: "You!"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='./')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes

---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
DOCNAME = gsub(DOCNAME, pattern = ".Rmd", replacement = "", fixed = T)
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("./img/", DOCNAME, "/"),
                      #fig.width      = 10,
                      #fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE,
                      eval           = FALSE)
```

Approximate time: 40 minutes

## Learning Objectives 

* Explore different types of normalization methods
* Become familiar with the `DESeqDataSet` object 
* Understand how to normalize counts using DESeq2

## Normalization

The first step in the DE analysis workflow is count normalization, which is necessary to make accurate comparisons of gene expression between samples.

***

**Exercise 1**

Determine the normalized (median of ratios) counts for your gene of interest, PD1, given the raw counts and size factors below. 

NOTE: You will need to run the code below to generate the raw counts dataframe (PD1) and the size factor vector (size_factors), then use these objects to determine the normalized counts values:

```{r}
# Raw counts for PD1
PD1 <- t(c(21, 58, 17, 97, 83, 10)) %>% 
  as_tibble() %>%
  rename_all(~paste0("Sample", 1:6))


# Size factors for each sample
size_factors <- c(1.32, 0.70, 1.04, 1.27, 1.11, 0.85)
```


```{r}
# your code here:

```

***

## Count normalization of Mov10 dataset using DESeq2

Now that we know the theory of count normalization, we will normalize the counts for the Mov10 dataset using DESeq2. This requires a few steps:

1. Ensure the row names of the metadata dataframe are present and in the same order as the column names of the counts dataframe.
2. Create a `DESeqDataSet` object
3. Generate the normalized counts

### 1. Match the metadata and counts data

We should always make sure that we have sample names that match between the two files, and that the samples are in the right order. DESeq2 will output an error if this is not the case.

```{r}
### Check that sample names match in both files
all(meta$samplename %in% colnames(data))
all(meta$samplename == colnames(data)[-1])
```

If your data did not match, you could use the `match()` function to rearrange them to be matching.

***

**Exercise 2**	

Suppose we had sample names matching in the counts matrix and metadata file, but they were out of order. Write the line(s) of code required to create a new matrix with columns ordered such that they were identical to the row names of the metadata. Remember that `data` contains the column `GeneSymbol` witht the names of the genes!

```{r}
# randomize count data columns and metadata rownames
meta_random <- meta[sample(1:nrow(meta)),]
data_random <- data[,c(1,sample(2:ncol(data)))]
```

```{r}
# your code here
```

*** 

### 2. Create DESEq2 object

Let's start by creating the `DESeqDataSet` object and then we can talk a bit more about what is stored inside it. To create the object we will need the **count matrix** and the **metadata** table as input. We will also need to specify a **design formula**. The design formula specifies the column(s) in the metadata table and how they should be used in the analysis. For our dataset we only have one column we are interested in, that is `~sampletype`. This column has three factor levels, which tells DESeq2 that for each gene we want to evaluate gene expression change with respect to these different levels.

```{r}
## Create DESeq2Dataset object
dds <- DESeqDataSetFromMatrix(countData = data.frame(data[,-1], row.names = data$GeneSymbol), 
                              colData = data.frame(meta[,-1], row.names = meta$samplename), 
                              design = ~ sampletype)
```

You can use DESeq-specific functions to access the different slots and retrieve information, if you wish. For example, suppose we wanted the original count matrix we would use `counts()` (*Note: we nested it within the `View()` function so that rather than getting printed in the console we can see it in the script editor*) :

```{r}
View(counts(dds))
```

As we go through the workflow we will use the relevant functions to check what information gets stored inside our object.

### 3. Generate the Mov10 normalized counts

The next step is to normalize the count data in order to be able to make fair gene comparisons between samples.

To perform the **median of ratios method** of normalization, DESeq2 has a single `estimateSizeFactors()` function that will generate size factors for us. We will use the function in the example below, but **in a typical RNA-seq analysis this step is automatically performed by the `DESeq()` function**, which we will see later. 

```{r}
dds <- estimateSizeFactors(dds)
```

By assigning the results back to the `dds` object we are filling in the slots of the `DESeqDataSet` object with the appropriate information. We can take a look at the normalization factor applied to each sample using:

```{r}
sizeFactors(dds)
```

Now, to retrieve the normalized counts matrix from `dds`, we use the `counts()` function and add the argument `normalized=TRUE`.

```{r}
normalized_counts <- counts(dds, normalized=TRUE)
```

We can save this normalized data matrix to file for later use:

```{r}
write.table(normalized_counts, file="/work/introduction_bulkRNAseq_analysis/Results/normalized_counts.txt", sep="\t", quote=F)
```
