---
title: "Genomic annotations for functional analyses"
author: "Jose Alejandro Romero Herrera"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='../develop/')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
  md_document:
    variant: gfm
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
DOCNAME = gsub(DOCNAME, pattern = ".Rmd", replacement = "", fixed = T)
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = TRUE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("./img/", DOCNAME, "/"),
                      #fig.width      = 10,
                      #fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE,
                      eval           = FALSE)
```

Approximate time: 30 minutes

## Learning Objectives 

* Discuss the available genomic annotation databases and the different types if information stored
* Compare and contrast the tools available for accessing genomic annotation databases 

## Genomic annotations

The analysis of next-generation sequencing results requires associating genes, transcripts, proteins, etc. with functional or regulatory information. To perform functional analysis on gene lists, we often need to obtain gene identifiers that are compatible with the tools we wish to use and this is not always trivial.

## Databases

We retrieve information on the processes, pathways, etc. (for which a gene is involved in) from the necessary database where the information is stored. The database you choose will be dependent on what type of information you are trying to obtain. Examples of databases that are often queried, include:

**General databases** 

Offer comprehensive information on genome features, feature coordinates, homology, variant information, phenotypes, protein domain/family information, associated biological processes/pathways, associated microRNAs, etc.:

- **Ensembl** (use Ensembl gene IDs)
- **NCBI** (use Entrez gene IDs)
- **UCSC** 
- **EMBL-EBI**

**Annotation-specific databases**

Provide annotations related to a specific topic:

- **Gene Ontology (GO):** database of gene ontology biological processes, cellular components and molecular functions - based on Ensembl or Entrez gene IDs or official gene symbols
- **KEGG:** database of biological pathways - based on Entrez gene IDs
- **MSigDB:** database of gene sets
- **Reactome:** database of biological pathways
- **Human Phenotype Ontology:** database of genes associated with human disease
- **CORUM:** database of protein complexes for human, mouse, rat
- **...**

## Tools for accessing databases

Within R, there are many popular packages used for gene/transcript-level annotation. These packages provide tools that take the list of genes you provide and retrieve information for each gene using one or more of the databases listed above.

### Annotation tools: for accessing/querying annotations from a specific databases

| Tool | Description | Pros | Cons |
|:---:|:---|:---|:---:|
|**[org.Xx.eg.db](https://bioconductor.org/packages/release/bioc/vignettes/AnnotationDbi/inst/doc/IntroToAnnotationPackages.pdf)** | Query gene feature information for the organism of interest | gene ID conversion, biotype and coordinate information | only latest genome build available |
|**[EnsDb.Xx.vxx](http://bioconductor.org/packages/devel/bioc/vignettes/ensembldb/inst/doc/ensembldb.html)**| Transcript and gene-level information directly fetched from Ensembl API (similar to TxDb, but with filtering ability and versioned by Ensembl release) | easy functions to extract features, direct filtering | Not the most up-to-date annotations, more difficult to use than some packages |
|**[TxDb.Xx.UCSC.hgxx.knownGene](https://bioconductor.org/packages/release/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.pdf)** | UCSC database for transcript and gene-level information or can create own *TxDb* from an SQLite database file using the *GenomicFeatures* package | feature information, easy functions to extract features | only available current and recent genome builds - can create your own, less up-to-date with annotations than Ensembl |
|**[annotables](https://github.com/stephenturner/annotables)** | Gene-level feature information immediately available for the human and model organisms | super quick and easy gene ID conversion, biotype and coordinate information | static resource, not updated regularly | 
|**[biomaRt](https://bioconductor.org/packages/release/bioc/vignettes/biomaRt/inst/doc/biomaRt.html)** | An R package version of the Ensembl [BioMart online tool](http://www.ensembl.org/biomart/martview/70dbbbe3f1c5389418b5ea1e02d89af3)  | all Ensembl database information available, all organisms on Ensembl, wealth of information |  |


## AnnotationHub

AnnotationHub is a wonderful resource for accessing genomic data or querying large collection of whole genome resources, including ENSEMBL, UCSC, ENCODE, Broad Institute, KEGG, NIH Pathway Interaction Database, etc. All of this information is stored and easily accessible by directly connecting to the database.

To get started with AnnotationHub, we first load the library and connect to the database:

```{r}
# Load libraries
library(AnnotationHub)
library(ensembldb)

# Connect to AnnotationHub
ah <- AnnotationHub()
```

To see the types of information stored inside our database, we can just type the name of the object:

```{r}
# Explore the AnnotationHub object
ah
```

  
If you would like to see more information about any of the classes of data you can extract that information as well. For example, if you wanted to **determine all species information available**, you could explore that within the AnnotationHub object:
  
```{r}
# Explore all species information available
unique(ah$species) %>% View()
```

In addition to species information, there is also additional information about the type of Data Objects and the Data Providers:
```{r}
# Explore the types of Data Objects available
unique(ah$rdataclass) %>% View()

# Explore the Data Providers
unique(ah$dataprovider) %>% View()
```
  
Now that we know the types of information available from AnnotationHub we can query it for the information we want using the `query()` function. Let's say we would like to **return the Ensembl `EnsDb` information for Human**. To return the records available, we need to use the terms as they are output from the `ah` object to extract the desired data.
  
```{r}
# Query AnnotationHub
human_ens <- query(ah, c("Homo sapiens", "EnsDb"))
```

The query retrieves all **hits for the `EnsDb` objects**, and you will see that they are listed by the release number. The most current release for GRCh38 is Ensembl98 and AnnotationHub offers that as an option to use. However, if you look at options for older releases, for Homo sapiens it only go back as far as Ensembl 87. This is fine if you are using GRCh38, however if you were using an older genome build like hg19/GRCh37, you would need to load the `EnsDb` package if available for that release or you might need to build your own with `ensembldb`.

```{r}
human_ens
```

If we are looking for the latest Ensembl release, so that the annotations are the most up-to-date, we can use the AnnotationHub ID to **subset the object**:

```{r}
# Extract annotations of interest
human_ens <- human_ens[["AH75011"]]
```

Now we can use `ensembldb` functions to extract the information at the gene, transcript, or exon levels. We are interested in the gene-level annotations, so we can extract that information as follows:

```{r}
# Extract gene-level information
genes(human_ens, return.type = "data.frame") %>% View()
```

But note that it is just as easy to get the transcript- or exon-level information:

```{r}
# Extract transcript-level information
transcripts(human_ens, return.type = "data.frame") %>% View()

# Extract exon-level information
exons(human_ens, return.type = "data.frame") %>% View()
```

To **obtain an annotation data frame** using AnnotationHub, we'll use the `genes()` function, but only keep selected columns and filter out rows to keep those corresponding to our gene identifiers in our results file:

```{r}
# Create a gene-level dataframe 
annotations_ahb <- genes(human_ens, return.type = "data.frame")  %>%
  dplyr::select(gene_id, gene_name, entrezid, gene_biotype, description) %>% 
  dplyr::filter(gene_name %in% res_tableOE_tb$gene)
```

This dataframe looks like it should be fine as it is, but we look a little closer we will notice that the column containing Entrez identifiers is a list, and in fact there are many Ensembl identifiers that map to more than one Entrez identifier!

```{r}
# Wait a second, we don't have one-to-one mappings!
class(annotations_ahb$entrezid)
which(map(annotations_ahb$entrezid, length) > 1)
```

So what do we do here? And why do we have this problem? An answer from the [Ensembl Help Desk](https://www.biostars.org/p/16505/) is that this occurs when we cannot choose a perfect match; ie when we have two good matches, but one does not appear to match with a better percentage than the other. In that case, we assign both matches. What we will do is choose to **keep the first identifier for these multiple mapping cases**.

```{r}
annotations_ahb$entrezid <- map(annotations_ahb$entrezid,1) %>%  unlist()
```

Let's take a look and see how many of our Ensembl identifiers have an associated gene symbol, and how many of them are unique:
```{r}
which(is.na(annotations_ahb$gene_name)) %>% length()

which(duplicated(annotations_ahb$gene_name)) %>% length()
```

Let's identify the non-duplicated genes and only keep the ones that are not duplicated:
```{r}
# Determine the indices for the non-duplicated genes
non_duplicates_idx <- which(duplicated(annotations_ahb$gene_name) == FALSE)

# How many rows does annotations_ahb have?
annotations_ahb %>% nrow()

# Return only the non-duplicated genes using indices
annotations_ahb <- annotations_ahb[non_duplicates_idx, ]

# How many rows are we left with after removing?
annotations_ahb %>% nrow()
```

Finally, it would be good to know **what proportion of the Ensembl identifiers map to an Entrez identifier**:

```{r}
# Determine how many of the Entrez column entries are NA
which(is.na(annotations_ahb$entrezid)) %>%  length()
```

For the different steps in the functional analysis, we require Ensembl and Entrez IDs. We will use the gene annotations that we generated previously to merge with our differential expression results.

```{r}
## Merge the AnnotationHub dataframe with the results 
res_ids_ahb <- left_join(res_tableOE_tb, annotations_ahb, by=c("gene"="gene_name"))    
```


### Using AnnotationHub to create a tx2gene file

If you wish to easily translate between genes and transcripts (this depends on how the count matrix was computed), it would be wise to create a 'transcript to gene' translation file (or `tx2gene` file). 
To create it, we would need to use a combination of the methods above and merge two dataframes together. For example:

```{r}
## DO NOT RUN THIS CODE

# Create a transcript dataframe
 txdb <- transcripts(human_ens, return.type = "data.frame") %>%
   dplyr::select(tx_id, gene_id)
 txdb <- txdb[grep("ENST", txdb$tx_id),]
 
 # Create a gene-level dataframe
 genedb <- genes(human_ens, return.type = "data.frame")  %>%
   dplyr::select(gene_id, gene_name)
 
 # Merge the two dataframes together
 annotations <- inner_join(txdb, genedb)
```

## Annotables package

The *annotables* package is a super easy annotation package to use. It is not updated frequently, so it's not great for getting the most up-to-date information for the current builds and does not have information for other organisms than human and mouse, but is a quick way to get annotation information, specially for older builds.

```{r}
library(annotables)
grch37
```

We can see that the `grch37` object already contains all the information we want in a super easy way. Let's annotate the results of our shrunken DEA for MOV10 overexpression:

```{r}
## Re-run this code if you are unsure that you have the right table
res_tableOE <- lfcShrink(dds, coef = "sampletype_MOV10_overexpression_vs_control")
res_tableOE_tb <- res_tableOE %>%
    data.frame() %>%
    rownames_to_column(var="gene") %>% 
    as_tibble()
```

```{r}
## Return the IDs for the gene symbols in the DE results
idx <- grch37$symbol %in% rownames(res_tableOE)

ids <- grch37[idx, ]

## The gene names can map to more than one Ensembl ID (some genes change ID over time), 
## so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)

ids <- ids[non_duplicates, ] 

## Merge the IDs with the results 
res_ids <- inner_join(res_tableOE_tb, ids, by=c("gene"="symbol"))
```

Our data is now ready to use for functional analysis! We have all the ids necessary to proceed.

***

**Exercise 1**

- Create a new `res_ids` object using the `annotables` package with the human build grch38. **NOTE** call it `res_ids_grch38`!
- What are the differences between the `res_id_ahb`object and the `res_ids_grch38`?

***